---
title: "Most Hazardous Weather Events in the U.S."
author: "Reynaldo Vazquez"
date: "9/07/2017"
output: html_document
---
<style>
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
  .col3 {
    columns: 3 100px;
    -webkit-columns: 3 100px;
    -moz-columns: 3 100px;
  }
  body {
  text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, cache = T, warning=F, message=F, 
                      fig.align = 'center')
```

## Synopsis  
The following analysis addresses two questions:  

1. In the U.S., what weather events have been the most harmful to population life and 
health?  
2. In the U.S., what weather events have been the most economically devastating?  

This analysis uses data from the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database. The database tracks characteristics of 
major storms and weather events in the United States, including date and 
location, estimates of fatalities, injuries, and property damage. The records 
start in 1950 and end in November 2011.  

Tornadoes are found to have caused the most fatalities and injuries in the 
period reported. The second most hazardous to human life and health, as well as 
the most economically devastating, are floods. 

## Data Processing

Data was downloaded from 
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) 
on September 7, 2017. The full raw dataset was loaded into R and processed by 
discarding variables not relevant to this analysis. Other transformations and 
manipulations are described throughout the document. 

1. Load required packages
```{r packages, warning=FALSE, message=FALSE}
require(dplyr); require(ggplot2);require(knitr); require(pander)
require(kableExtra);require(gridExtra);require(reshape2)
```
2. Load full dataset
```{r load, cache = T}
fulldata <- read.csv("repdata-data-StormData.csv.bz2", stringsAsFactors = FALSE)
```
3. Inspect variable names to discard those non-relevant 
```{r names}
names(fulldata)
```
4. Select needed variables and analyze their structure
```{r trim}
data <- select(fulldata, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, 
               CROPDMG, CROPDMGEXP)
str(data)
```
5. Preprocess event type names  

The original data set has `r length(unique(fulldata$EVTYPE))` *unique* types of 
events. Many of these seem to belong to a same category of event as others, as 
can be seen in the example below for hurricanes.
```{r example}
exIdx <- grep("HURR", data$EVTYPE, perl = TRUE)
unique(data$EVTYPE[exIdx])
```

The following preprocessing partially solves the issue above by correcting some 
misspellings, substituting plurals and synonyms, removing excessive space, 
symbols, numbers, redundancies, and specificities. The purpose is to rename 
original *event types* that unambiguously belong to the same type. 

```{r grepPrep}
remove <- grep("summary", tolower(data$EVTYPE), perl=TRUE)
data   <- data[-remove,]
events <- tolower(data$EVTYPE)
events <- gsub("[[:alnum:]]*[.]*[0-9]+", "", events, perl=TRUE, useBytes = T)
events <- gsub("[^[:alpha:][:space:]]|mph", " ", events, perl=T, useBytes = T)
events <- gsub("(?<=hurricane)[ a-z]*", "", events, perl=TRUE, useBytes = T)
events <- gsub("(?<=tropical storm)[ a-z]*", "", events, perl=TRUE, useBytes = T)
events <- gsub("thunde[a-z]*", "thunderstorm", events, perl=TRUE, useBytes = T)
events <- gsub("tstm", "thunderstorm", events, perl=TRUE, useBytes = T)
events <- gsub("floo[a-z]*", "floods", events, perl=TRUE, useBytes = T)
events <- gsub("precipit[a-z]*|rain[a-z]*", "rain", events, perl=TRUE, useBytes = T)
events <- gsub("winds", "wind", events, perl=TRUE, useBytes = T)
events <- gsub("torn[a-z]+", "tornado", events, perl=TRUE, useBytes = T)
events <- gsub("thunde[a-z]*", "thunderstorm", events, perl=TRUE, useBytes = T)
pattrn <- "extr[a-z]*|record|severe|exce[a-z]*|unusu[a-z]*|heavy|major|abno[a-z]*|haza[a-z]*"
events <- gsub(pattrn, "excessive", events, perl=TRUE, useBytes = T)
pattrn <- "astron[a-z]*|weather|temperat[a-z]*|damag[a-z]*"
events <- gsub(pattrn, "", events, perl=TRUE, useBytes = T)
events <- gsub("(?<=[\\s])\\s*|^\\s+|\\s+$", "", events, perl=TRUE, useBytes = T)
reCap  <- function(text) {
    s  <- strsplit(text, " ")[[1]]
    paste(toupper(substring(s, 1,1)), substring(s, 2), sep="", collapse=" ")
}
events <- sapply(events, reCap, USE.NAMES = FALSE)
data$EVTYPE <- events
```

The event type preprocessing above reduced the number of *unique* event 
categories from `r length(unique(fulldata$EVTYPE))` to `r length(unique(events))`. 
Although this recategorization is not comprehensive, it should suffice for the 
rest of the analysis. Skipping this recategorization has only limited 
quantitative effects on the results, whereas the qualitative effects are 
negligible, at least for the most devastating types of events. 

6. Decode PROPDMGEXP and CROPDMGEXP, with their corresponding values. These two 
variables contain either: a number, or a letter representing the exponent that 
multiplies their corresponding numeric values in the PROPDM and CROPDMG columns
```{r exps}
data$PROPDMGEXP <- tolower(data$PROPDMGEXP)
data$CROPDMGEXP <- tolower(data$CROPDMGEXP)
xfactors <- c("h", "k", "m", "b")
xvalues  <- c("2", "3", "6", "9")
for (i in 1:4){
        data$PROPDMGEXP <- gsub(xfactors[i], xvalues[i], data$PROPDMGEXP, fixed = TRUE)
        data$CROPDMGEXP <- gsub(xfactors[i], xvalues[i], data$CROPDMGEXP, fixed = TRUE)
}
```

7. Transform the PROPDMG and CROPDMG to contain the dollar amounts of their 
respective damages. Calculate a new variable "TotalDamage", by adding the two 
types of damage amounts. 
```{r damages, warning=FALSE, message=FALSE}
data$PROPDMGEXP <- 10^as.numeric(as.character(data$PROPDMGEXP))
data$CROPDMGEXP <- 10^as.numeric(as.character(data$CROPDMGEXP))
data$PROPDMG    <- data$PROPDMG*data$PROPDMGEXP
data$CROPDMG    <- data$CROPDMG*data$CROPDMGEXP
TotalDamage     <- rowSums(cbind(data$PROPDMG, data$CROPDMG), na.rm = TRUE)
```

8. Tidy up the data by keeping only variables needed and fixing variable names
```{r tidy}
data <- select(data, EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG)
names(data) <- c("EventType", "Fatalities", "Injuries", "Property", "Crops")
data <- cbind(data, TotalDamage)
```

9. Calculate the sum of total fatalities and injuries by type of event
```{r sums}
sums <- aggregate(cbind(Fatalities, Injuries) ~ EventType, data = data, sum)
```
   
10. Select and rank the 10 weather events with most deaths and most injuries, 
respectively
```{r tables}
tops <- 10
TF   <- head(select(arrange(sums, desc(Fatalities)), EventType, Fatalities), tops)
Rank <- c(1:tops)
TF2  <- cbind(Rank, TF)
TF2[,3] <- format(TF2[,3], decimal.mark=".", big.mark=",", scientific=F, digits = 0)
TI   <- head(select(arrange(sums, desc(Injuries)), EventType, Injuries), tops)
TI2  <- cbind(Rank, TI)
TI2[,3] <- format(TI2[,3], decimal.mark=".", big.mark=",", scientific=F, digits = 0)
deathsTable <- kable(TF2, format = "html", caption = "Table 1: Deadliest Weather Events",
                align=c("c", "l", "r"), col.names = c("Rank", "Type of Event", "Fatalities")) %>%
    kable_styling(bootstrap_options = c("striped","condensed", "bordered"), full_width = F) %>%
    add_footnote(c("period 1950 - 2011"), notation = "alphabet")
injuriesTable <- kable(TI2, format = "html", caption = "Table 2: Events with Most Injuries",
                  align=c("c", "l", "r"), col.names = c("Rank", "Type of Event", "Injuries")) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "bordered"), full_width = F) %>%
    add_footnote(c("period 1950 - 2011"), notation = "alphabet")
```

<div class="col2">
```{r tables2}
deathsTable
```

```{r tables3}
injuriesTable
```
</div>

11. Create graphs showing the events with most deaths and most injuries, 
respectively
```{r graphs1, fig.width=9.5, fig.height=4}
TF$EventType <- factor(TF$EventType, levels = TF$EventType)
g1 <- ggplot(TF, aes(EventType, Fatalities)) +
        geom_bar(stat = "identity", fill = "red4", alpha = 0.9) + theme_bw() +
        theme(axis.text.x = element_text(angle = 25, hjust = 1)) + xlab("Event Type") + 
        ggtitle("Fig. 1: Deadliest Weather Events\n         (1950 - 2011 Total)")
TI$EventType <- factor(TI$EventType, levels = TI$EventType)
g2 <- ggplot(TI, aes(EventType, Injuries)) + 
        geom_bar(stat = "identity", fill = "steelblue", alpha = 0.9) +theme_bw() +
        theme(axis.text.x = element_text(angle = 25, hjust = 1)) + xlab("Event Type") + 
        ggtitle("Fig. 2: Weather Events with Most Injuries\n         (1950 - 2011 Total)")
grid.arrange(g1, g2, ncol = 2)
```
  
12. Calculate the sum of total damage, property and crops costs by type of 
event. Then select and rank the 10 costliest weather events
```{r Damages}
sumsD <- aggregate(x = list(TotalDamage = data$TotalDamage, 
                            Property = data$Property, 
                            Crops = data$Crops), 
                   by = list(EventType = data$EventType), FUN = sum, 
        na.rm = TRUE)
TD    <- head(select(arrange(sumsD, desc(TotalDamage)), EventType, 
                     TotalDamage, Property, Crops), 10)
```

13. Transform damage values to USD Millions for illustration. Prepare and print 
table
```{r Damages2}
TDD <- TD
TDD[,2:4] <- round(TDD[,2:4]/10^6, digits = 3)
TDD[,2:4] <- format(TDD[,2:4], decimal.mark=".", big.mark=",", scientific=F, digits = 0)
TDD <- cbind(TDD[,-2], TDD[,2])
TDD <- cbind(Rank, TDD)
kable(TDD, format = "html", caption = "Table 1: Costliest Weather Events (USD Millions)",
                    align=c("c", "l", "r", "r", "r"), 
                    col.names = c("Rank", "Type of Event", "Damage to Property", 
                                  "Damage to Crops", "Total Damage")) %>%
    kable_styling(bootstrap_options = c("striped","condensed", "bordered"), full_width = F) %>%
    add_footnote(c("period 1950 - 2011"), notation = "alphabet")
```

14. Create graph showing costliest events
```{r Damages4}
TD$EventType <- factor(TD$EventType, levels = TD$EventType)
TD2 <- melt(TD[,c(1,4,3)], id.vars = "EventType")
ggplot(TD2, aes(x = EventType, y = value/10^9, fill = variable)) + 
        geom_bar(stat = 'identity', alpha = 0.9) + theme_bw() +
        theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
        xlab("Event Type") + ylab("Economic Damages (USD Bilions)") +
        ggtitle("Fig. 3: Costliest Weather Events (USD Billions)") + 
        scale_fill_manual(values=c("red4", "steelblue")) +
        theme(legend.title = element_blank())
```

## Results   

The main results are summarized in Tables 1, 2, and 3 and their corresponding 
figures. The data analyzed shows tornadoes are the events that have caused the 
most fatalities and injuries in the population in the years reported. The second 
most hazardous to human life and health, as well as the most economically 
devastating weather type of events are floods.   

  
